### 2025-07-21 paper/quarto_aaai/paper.tex
done with editorail pass

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} follows \citet{altmeyer2024faithful}, with
adapted notation. Intuitively, plausible counterfactuals are consistent
with the data and faithful counterfactuals are consistent with what the
model has learned about the input data. Ac tionability constraints in
Def. \ref{def-explainability} vary and depend on the context in which
\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
consider mutability constraints for individual features \(x_d\) for
\(d=1,...,D\). We also limit ourselves to classification tasks for
reasons discussed in Section~\ref{sec-discussion}.

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any explainer that relies on gradient descent in the
feature space for optimization \citep{altmeyer2023explaining}. In this
context, domain constraints can be imposed by simply projecting
counterfactuals back to the specified domain, if the previous gradient
step resulted in updated feature values that were out-of-domain.
Similarly, mutability constraints can be enforced by setting partial
derivatives to zero to ensure that features are only perturbed in the
allowed direction, if at all.

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to inform model representations. Through their effect on
\(\mathbf{x}^\prime\), both types of constraints influence model
outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
penalizing implausibility that arises due to mutability constraints. For
any mutability-constrained feature \(d\) this can be achieved by
enforcing \(\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0\) whenever
perturbing \(\mathbf{x}^\prime[d]\) in the direction of
\(\mathbf{x}^+[d]\) would violate mutability constraints. Specifically,
we set \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) if:

@@

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) wrt. feature \(d\).
Intuitively, we think of this as ignoring implausibility costs of
immutable features, which effectively forces the model to instead seek
plausibility through the remaining features. This can be expected to
result in lower overall sensitivity to immutable features, which we
investigate empirically in Section~\ref{sec-experiments}. Under certain
conditions, this result holds theoretically; for the proof, see the
supplementary appendix:

@@

Our focus is the improvement in explainability (Def.
\ref{def-explainability}). Thus, we primarily look at the plausibility
and cost of faithfully generated counterfactuals at test time. Other
metrics, such as validity and redundancy, are reported in the
supplementary appendix. To measure the cost, we follow the standard
proxy of distances (\(\ell_1\)-norm) between factuals and
counterfactuals. For plausibility, we assess how similar CEs are to the
observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence-based adaptation,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). They are generated using the library of
\citet{altmeyer2023explaining} and we present them in the supplementary
appendix. Next, we have four real-world binary tabular datasets:
\emph{Adult} (Census data) of \citet{becker1996adult2}, California
housing (\emph{CH}) of \citet{pace1997sparse}, Default of Credit Card
Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me Some Credit
(\emph{GMSC}) from \citet{kaggle2011give}. Finally, for the convenience
of illustration, we use the 10-class \emph{MNIST}
\citep{lecun1998mnist}.

@@

The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
percentage reduction in implausibility for our two metrics when using CT
on top of the weak baseline. As an example, consider the first row for
\emph{LS} data: the observed positive values indicate that faithful
counterfactuals are around 30-55\% more plausible for models trained
with CT, in line with our observations in Figure~\ref{fig-poc} in panel
(b) compared to panel (a).

@@

\emph{CT generally produces substantial and statistically significant
improvements in plausibility.}\\
Average reductions in \(\text{IP}\) range from around 7\% for
\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
are even more substantial and generally statistically significant,
although the average degree of uncertainty is higher than for
\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
almost 90\% (\emph{Circ}). The only negative findings are for OL and
MNIST, but they are not statistically significant. A qualitative
inspection of the counterfactuals in Figure~\ref{fig-mnist} (columns
2-5) suggests recognizable digits 1-4 for the model trained with CT
(bottom row), unlike the baseline (top row).

@@

\emph{CT tends to improve actionability in the presence of immutable
features, but the exact effect is sometimes hard to establish due to,
e.g., feature interactions.}

@@

First, we look at hyperparameters that guide the gradient-based
counterfactual search. Increasing \(T\), the maximum number of steps,
generally yields better outcomes because more CEs can mature in each
training epoch. The impact of \(\tau\), the required decision threshold
is more difficult to predict. On ``harder'' datasets it may be difficult
to satisfy high \(\tau\) for any given sample (i.e., also factuals) and
so increasing this threshold does not seem to correlate with better
outcomes. In fact, \(\tau=0.5\) generally leads to optimal results as it
is associated with high proportions of mature counterfactuals.

Second, the strength of the energy regularization,
\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
performance in terms of decreased plausibility and increased costs if
insufficiently high. The sensitivity with respect to
\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
evident. While high values of \(\lambda_{\text{reg}}\) may increase the
variability in outcomes when combined with high values of
\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
particularly pronounced.

Third, the effectiveness and stability of CT is positively associated
with the number of counterfactuals generated during each training epoch.
A higher number of training epochs is also beneficial. Interestingly, we
observed desired improvements when CT was combined with conventional
training and applied only for the final 50\% of epochs of the complete
training process. Put differently, CT can improve the explainability of
models in a fine-tuning manner.

@@

\emph{CT increases the training times.} Like adversarial training, CT is
more resource-intensive than conventional regimes. Higher numbers of CEs
improve the quality of the learned representations but they also
increase the number of computations. As our codebase is not performance
optimized, grids of around 300 settings for the largest datasets in our
experiments took up to four hours using 34 2GB CPUs (see supplementary
appendix). Other than optimization, three factors mitigate this effect:
(1) CT yields itself to parallel execution; (2) it amortizes the cost of
CEs for the training samples; and (3) it can be used to fine-tune
conventionally-trained models.

@@

Secondly, CT is susceptible to training instabilities, as is also the
case for the related JEMs \citep{grathwohl2020your}. CT is exposed to
two sources of instabilities: (1) the energy-based contrastive
divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
(2) the underlying explainers. We find several promising ways to
mitigate this problem: regularizing energy (\(\lambda_{\text{reg}}\)),
generating sufficiently many counterfactuals during each epoch, and
including only mature counterfactuals in \(\text{div}(\cdot)\).

>>>>

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} follows \citet{altmeyer2024faithful}, with
adapted notation. Intuitively, plausible counterfactuals are consistent
with the data and faithful counterfactuals are consistent with what the
model has learned about the input data. Ac tionability constraints in
Def. \ref{def-explainability} vary and depend on the context in which
\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
consider domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We also limit ourselves to classification
tasks for reasons discussed in Section~\ref{sec-discussion}.

@@

Many existing counterfactual explainers support domain and mutability
constraints. In fact, both types of constraints can be implemented for
any explainer that relies on gradient descent in the feature space for
optimization \citep{altmeyer2023explaining}. In this context, domain
constraints can be imposed by simply projecting counterfactuals back to
the specified domain, if the previous gradient step resulted in updated
feature values that were out-of-domain. Similarly, mutability
constraints can be enforced by setting partial derivatives to zero to
ensure that features are only perturbed in the allowed direction, if at
all.

Since actionability constraints are binding at test time, we also impose
them when generating \(\mathbf{x}^\prime\) during each training
iteration to inform model representations. Through their effect on
\(\mathbf{x}^\prime\), both types of constraints influence model
outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
penalizing implausibility that arises due to mutability constraints. For
any mutability-constrained feature \(d\) this can be achieved by
enforcing \(\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0\) whenever
perturbing \(\mathbf{x}^\prime[d]\) in the direction of
\(\mathbf{x}^+[d]\) would violate mutability constraints. Specifically,
we set \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) if:

@@

\noindent From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) wrt. feature \(d\).
Intuitively, we think of this as ignoring implausibility costs of
immutable features, which effectively forces the model to instead seek
plausibility through the remaining features. This can be expected to
result in relatively lower sensitivity to immutable features; and higher
relative sensitivity to mutable features should make
mutability-constrained recourse less costly
(Section~\ref{sec-experiments}). Under certain conditions, this result
holds theoretically; for the proof, see the supplementary appendix:

@@

Our focus is the improvement in explainability (Def.
\ref{def-explainability}). Thus, we primarily look at the plausibility
and cost of faithfully generated counterfactuals at test time. Other
metrics, such as validity and redundancy, are reported in the
supplementary appendix. To measure the cost, we follow the standard
proxy of distances (\(\ell_1\)-norm) between factuals and
counterfactuals. For plausibility, we assess how similar CEs are to
observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence-based adaptation,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). Next, we have four real-world binary tabular
datasets: \emph{Adult} (Census data) of \citet{becker1996adult2},
California housing (\emph{CH}) of \citet{pace1997sparse}, Default of
Credit Card Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me
Some Credit (\emph{GMSC}) from \citet{kaggle2011give}. Finally, for the
convenience of illustration, we use the 10-class \emph{MNIST}
\citep{lecun1998mnist}.

@@

The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
percentage reduction in implausibility for our two metrics when using CT
on top of the weak baseline. As an example, consider the first row for
\emph{LS} data: the observed positive values indicate that faithful
counterfactuals are around 30-55\% more plausible for models trained
with CT, in line with our observations in panel (b) of
Figure~\ref{fig-poc} compared to panel (a).

@@

\emph{CT generally produces substantial and statistically significant
improvements in plausibility.}

Average reductions in \(\text{IP}\) range from around 7\% for
\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
are even more substantial and generally statistically significant,
although the average degree of uncertainty is higher than for
\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
almost 90\% (\emph{Circ}). The only negative findings are for OL and
MNIST, but they are not statistically significant. A qualitative
inspection of the counterfactuals in Figure~\ref{fig-mnist} (columns
2-5) suggests recognizable digits 1-4 for the model trained with CT
(bottom row), unlike the baseline (top row).

@@

\emph{CT tends to improve actionability in the presence of immutable
features, but this is not guaranteed if the assumptions in Prp.
\ref{prp-mtblty} are violated.}

@@

Firstly, we find that optimal results are generally obtained when using
\emph{ECCCo} to generate counterfactuals. Conversely, using a generator
that may inhibit faithfulness (\emph{REVISE}), tends to yield poor
results. Concerning hyperparameters that guide the gradient-based
counterfactual search, we find that increasing \(T\), the maximum number
of steps, generally yields better outcomes because more CEs can mature.
Relatedly, we also find that the effectiveness and stability of CT is
positively associated with the total number of counterfactuals generated
during each training epoch. The impact of \(\tau\), the decision
threshold, is more difficult to predict. On ``harder'' datasets it may
be difficult to satisfy high \(\tau\) for any given sample (i.e., also
factuals) and so increasing this threshold does not seem to correlate
with better outcomes. In fact, \(\tau=0.5\) generally leads to optimal
results as it is associated with high proportions of mature
counterfactuals.

Secondly, the strength of the energy regularization,
\(\lambda_{\text{reg}}\) is highly impactful and should be set
sufficiently high to avoid common problems associated with exploding
gradients. The sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values of \(\lambda_{\text{div}}\) or
\(\lambda_{\text{adv}}\), this effect is not particularly pronounced.

Finally, we also observe desired improvements when CT was combined with
conventional training and applied only for the final 50\% of epochs of
the complete training process. Put differently, CT can improve the
explainability of models in a post-hoc, fine-tuning manner.

@@

\emph{Plausibility is costly.} As noted elsewhere
\citep{altmeyer2024faithful}, more plausible counterfactuals are
inevitably more costly. Thus, while CT improves plausibility and
robustness, it can impact average costs and validity as cheap,
implausible and adversarial explanations are removed from the solution
space.

\emph{CT increases the training times.} Like contrastive and robust
learning, CT is more resource-intensive than conventional regimes. Three
factors mitigate this effect: (1) CT yields itself to parallel
execution; (2) it amortizes the cost of CEs for the training samples;
and (3) it can be used to fine-tune conventionally-trained models.

@@

Secondly, it is worth extending the analysis done here to additional CE
generators including those that do not rely on computationally costly
gradient-based optimization. This would reduce training costs while
potentially maintaining benefits of CT.

### 2025-07-21 paper/sections/discussion.qmd
done with editorail pass

*CT increases the training times.* Like adversarial training, CT is more resource-intensive than conventional regimes. Higher numbers of CEs improve the quality of the learned representations but they also increase the number of computations. As our codebase is not performance optimized, grids of around 300 settings for the largest datasets in our experiments took up to four hours using 34 2GB CPUs (see supplementary appendix). Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.

We also highlight three key directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) and target class(es), requiring the output space to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR; other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.  
Secondly, CT is susceptible to training instabilities, as also recognized for the related JEMs [@grathwohl2020your]. CT is exposed to two sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.  
Finally, we believe that it is possible to considerably improve hyperparameter selection procedures, and thus performance. We have relied exclusively on grid searches, but future work could benefit from more sophisticated approaches.

>>>>

*Plausibility is costly.* As noted elsewhere [@altmeyer2024faithful], more plausible counterfactuals are inevitably more costly. Thus, while CT improves plausibility and robustness, it can impact average costs and validity as cheap, implausible and adversarial explanations are removed from the solution space.

*CT increases the training times.* Like contrastive and robust learning, CT is more resource-intensive than conventional regimes. Three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.

We also highlight three key directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) and target class(es), requiring the output space to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized. Focus on classification is a common choice in research on CEs and AR; other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.

Secondly, it is worth extending the analysis done here to additional CE generators including those that do not rely on computationally costly gradient-based optimization. This would reduce training costs while potentially maintaining benefits of CT.

Finally, we believe that it is possible to considerably improve hyperparameter selection procedures, and thus performance. We have relied exclusively on grid searches, but future work could benefit from more sophisticated approaches.

### 2025-07-21 paper/sections/experiments.qmd
done with editorail pass

Our focus is the improvement in explainability (Def. \ref{def-explainability}). Thus, we primarily look at the plausibility and cost of faithfully generated counterfactuals at test time. Other metrics, such as validity and redundancy, are reported in the supplementary appendix. To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence-based adaptation,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

@@

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). They are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix. Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give. Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].

@@

The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline. As an example, consider the first row for *LS* data: the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc in panel (b) compared to panel (a).

@@

### Plausibility (RQ1). {#sec-plaus}
*CT generally produces substantial and statistically significant improvements in plausibility.*  
Average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*. For the real-world tabular datasets they are around 12% for *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$. Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings are for OL and MNIST, but they are not statistically significant. A qualitative inspection of the counterfactuals in @fig-mnist (columns 2-5) suggests recognizable digits 1-4 for the model trained with CT (bottom row), unlike the baseline (top row).

@@

### Actionability (RQ2). {#sec-act}
*CT tends to improve actionability in the presence of immutable features, but the exact effect is sometimes hard to establish due to, e.g., feature interactions.*

@@

First, we look at hyperparameters that guide the gradient-based counterfactual search. Increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. The impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, $\tau=0.5$ generally leads to optimal results as it is associated with high proportions of mature counterfactuals.

Second, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not particularly pronounced.

Third, the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. A higher number of training epochs is also beneficial. Interestingly, we observed desired improvements when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT can improve the explainability of models in a fine-tuning manner.

>>>>

Our focus is the improvement in explainability (Def. \ref{def-explainability}). Thus, we primarily look at the plausibility and cost of faithfully generated counterfactuals at test time. Other metrics, such as validity and redundancy, are reported in the supplementary appendix. To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals. For plausibility, we assess how similar CEs are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence-based adaptation,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

@@

We make use of nine classification datasets common in the CE/AR literature. Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*). Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give. Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].

@@

The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline. As an example, consider the first row for *LS* data: the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in panel (b) of @fig-poc compared to panel (a).

@@

### Plausibility (RQ1). {#sec-plaus}
*CT generally produces substantial and statistically significant improvements in plausibility.*

Average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*. For the real-world tabular datasets they are around 12% for *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$. Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings are for OL and MNIST, but they are not statistically significant. A qualitative inspection of the counterfactuals in @fig-mnist (columns 2-5) suggests recognizable digits 1-4 for the model trained with CT (bottom row), unlike the baseline (top row).

@@

### Actionability (RQ2). {#sec-act}
*CT tends to improve actionability in the presence of immutable features, but this is not guaranteed if the assumptions in Prp. \ref{prp-mtblty} are violated.*

@@

Firstly, we find that optimal results are generally obtained when using *ECCCo* to generate counterfactuals. Conversely, using a generator that may inhibit faithfulness (*REVISE*), tends to yield poor results. Concerning hyperparameters that guide the gradient-based counterfactual search, we find that increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature. Relatedly, we also find that the effectiveness and stability of CT is positively associated with the total number of counterfactuals generated during each training epoch. The impact of $\tau$, the decision threshold, is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, $\tau=0.5$ generally leads to optimal results as it is associated with high proportions of mature counterfactuals.

Secondly, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and should be set sufficiently high to avoid common problems associated with exploding gradients. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not particularly pronounced.

Finally, we also observe desired improvements when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT can improve the explainability of models in a post-hoc, fine-tuning manner.

### 2025-07-21 paper/sections/method.qmd
done with editorail pass

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} follows @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Ac tionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

Many existing counterfactual explainers support domain and mutability constraints out-of-the-box. In fact, both types of constraints can be implemented for any explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining]. In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain. Similarly, mutability constraints can be enforced by setting partial derivatives to zero to ensure that features are only perturbed in the allowed direction, if at all.

Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}^+[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ if:

@@

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ wrt. feature $d$. Intuitively, we think of this as ignoring implausibility costs of immutable features, which effectively forces the model to instead seek plausibility through the remaining features. This can be expected to result in lower overall sensitivity to immutable features, which we investigate empirically in @sec-experiments. Under certain conditions, this result holds theoretically; for the proof, see the supplementary appendix:

>>>>

The characterization of faithfulness and plausibility in Def. \ref{def-explainability} follows @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Ac tionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

Many existing counterfactual explainers support domain and mutability constraints. In fact, both types of constraints can be implemented for any explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining]. In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain. Similarly, mutability constraints can be enforced by setting partial derivatives to zero to ensure that features are only perturbed in the allowed direction, if at all.

Since actionability constraints are binding at test time, we also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}^+[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ if:

@@

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ wrt. feature $d$. Intuitively, we think of this as ignoring implausibility costs of immutable features, which effectively forces the model to instead seek plausibility through the remaining features. This can be expected to result in relatively lower sensitivity to immutable features; and higher relative sensitivity to mutable features should make mutability-constrained recourse less costly (@sec-experiments). Under certain conditions, this result holds theoretically; for the proof, see the supplementary appendix:
